### 2025-04-14 Thesis/Latex/Chapter_3.tex
update

Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications. Furthermore, it is possible to \textbf{decompose any high-rank matrix} into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.
\subsection{Algorithm Design}

we introduce an adaptive threshold mechanism
(Dynamic Candidate Selection), the MoL framework is employed to refine the
candidate retrieval process. 
1. \textbf{Component-Level Embeddings}:  
Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
\begin{equation}
	X_p \leftarrow \{g_p(x) \mid x \in X\}.
\end{equation}

\noindent
2. \textbf{Dynamic Threshold Adjustment}:  
To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\). The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores. The scoring function is defined as:
\begin{equation}
	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\end{equation}
The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
\begin{equation}
	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
\end{equation}

\noindent
3. \textbf{Refinement and Top-K Selection}:  
Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\). This is achieved by including candidates whose scores exceed the threshold:
\begin{equation}
	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
\end{equation}
The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.

\noindent
4. \textbf{Exact Top-K Selection}:  
Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
\begin{equation}
	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
\end{equation}

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
	\caption{propused solution steps}
	\label{propused _solution_steps}	
\end{figure}

@@

\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

@@

	\textbf{1. Initial Candidate Retrieval:} \\
	\ForEach{component $p \in P$}{
		Compute dot product scores: 
		\[
		S_p \gets \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
		\]
		Retrieve items with scores $S_p \geq T_{\text{init}}$ \\
		Add these items to $G$
	}

	\textbf{2. Adjust k Dynamically:} \\
	\ForEach{$x \in G$}{
		Compute MoL scores $s \gets \phi(q, x)$ where:
		\[
		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
		\]
		Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
	}

@@

	\textbf{Return:} $G_{\text{final}}$ \tcp{Retrieve Top $k$ items from $G'$}
\end{algorithm}
\section{Recommendation System Overview}
Modern technology and online services have enabled unprecedented access to vast amounts of data. However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently. Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}. These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
	\caption{Recommendation System Process}
	\label{Recommendation_System _Process}	
\end{figure}
\subsection{SASRec: Self-Attentive Sequential Recommendation}
Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions. However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
\subsection{SASRec Model Architecture}
SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
	\caption{the training process of SASRec}
	\label{the_training_process_of_SASRec}	
\end{figure}
\section{Evaluation Metrics}
We employ standard ranking metrics widely used in sequential recommendation:

>>>>

Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications. Furthermore, it is possible to \textbf{decompose any high-rank matrix} into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.
\subsection{Algorithm Design}
we introduce an adaptive threshold mechanism
(Dynamic Candidate Selection), the MoL framework is employed to refine the
candidate retrieval process. \\
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
	\caption{propused solution steps}
	\label{propused _solution_steps}	
\end{figure}
\noindent
1. \textbf{Component-Level Embeddings}:  
Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
\begin{equation}
	X_p \leftarrow \{g_p(x) \mid x \in X\}.
\end{equation}

\noindent
2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.

\begin{equation}
	S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
	\label{eq:initial_retrieval}
\end{equation}
Here, $f_p(q)$ and $g_p(x)$ represent the feature embeddings of the query $q$ and item $x$ for the $p$-th component, respectively. The dot product $\langle f_p(q), g_p(x) \rangle$ measures the relevance of each item $x \in X_p$ with respect to $q$, producing a set of scores $S_p$ .

\noindent
3. \textbf{Dynamic Threshold Adjustment}:  
To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\). The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores. The scoring function is defined as:
\begin{equation}
	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
	\label{similarity_function}
\end{equation}
The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
\begin{equation}
	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
\end{equation}

\noindent
4. \textbf{Refinement and Top-K Selection}:  
Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\). This is achieved by including candidates whose scores exceed the threshold:
\begin{equation}
	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
\end{equation}
The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.

\noindent
4. \textbf{Exact Top-K Selection}:  
Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
\begin{equation}
	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
\end{equation}
As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

@@

	\textbf{1. Initial Candidate Retrieval:} \\
	\ForEach{component $p \in P$}{
		Compute dot product scores: 
		with (eq\eqref{eq:initial_retrieval}
		Retrieve items with scores $S_p \geq T_{\text{init}}$ \\
		Add these items to $G$
	}

	\textbf{2. Adjust k Dynamically:} \\
	\ForEach{$x \in G$}{
		Compute MoL scores $s \gets \phi(q, x)$  using : eq\eqref{similarity_function}
		Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
	}

@@

	\textbf{Return:} $G_{\text{final}}$ \tcp{Retrieve Top $k$ items from $G'$}
\end{algorithm}

\newpage

\section{Recommendation System Overview}
Modern technology and online services have enabled unprecedented access to vast amounts of data. However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently. Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}. These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
	\caption{Recommendation System Process}
	\label{Recommendation_System _Process}	
\end{figure}
\subsection{SASRec: Self-Attentive Sequential Recommendation}
Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions. However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
\subsection{SASRec Model Architecture}
SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
	\caption{the training process of SASRec}
	\label{the_training_process_of_SASRec}	
\end{figure}
\section{Evaluation Metrics}
We employ standard ranking metrics widely used in sequential recommendation:

### 2025-04-14 Thesis/Latex/General introduction.tex
update

\chapter*{General introduction}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{General introduction}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-MascaraComputer Science: 2025}}}
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
\pagenumbering{arabic}
\addcontentsline{toc}{chapter}{General introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Context :}
In recent years, the legal domain has witnessed a growing interest in the application of advanced natural language processing techniques to improve access to legal information and automate reasoning over juridical data. Legal texts—such as statutes, court decisions, and regulatory documents—are often dense, context-dependent, and linguistically complex, making them challenging to interpret both for non-experts and legal professionals. This complexity is further amplified in multilingual and region-specific legal systems, such as Algeria's, where access to structured and searchable legal data remains limited.

Large Language Models (LLMs) have emerged as powerful tools for understanding and generating human-like text, showing promise in tasks such as legal question answering, case summarization, and document classification. However, their reliance on static training data and the inherent difficulty in capturing fine-grained, up-to-date legal knowledge limits their effectiveness in dynamic and high-stakes domains like law. To bridge this gap, Retrieval-Augmented Generation (RAG) architectures have been proposed, combining the linguistic capabilities of LLMs with external knowledge retrieval systems. This hybrid approach enables models to access and reason over relevant documents at inference time, offering a more grounded and contextually aware output.

Despite these advances, several challenges persist—particularly in the selection of the most relevant documents from large legal corpora. Poor retrieval quality can significantly degrade the performance and reliability of RAG-based systems. Therefore, refining the retrieval process remains a crucial area of research, especially in low-resource and domain-specific contexts such as Arabic legal data in Algeria.\\
\textbf{Problem Statement :} 
While LLMs and RAG architectures provide a strong foundation for legal AI applications, their effectiveness is often undermined by limitations in the retrieval process. The selection of irrelevant or marginally related documents can lead to hallucinated or misleading responses, which is especially problematic in legal reasoning tasks. In the Algerian legal context, where structured resources are scarce and many legal documents exist only in unstructured Arabic text, the challenge of effective document retrieval becomes even more critical. This thesis addresses the specific problem of optimizing the retrieval step in RAG pipelines by introducing a more adaptive and context-sensitive document selection mechanism tailored to the legal domain.\\
\textbf{Objectives:} 
This research aims to explore and improve the integration of LLMs and retrieval systems in the legal field, with a focus on the Algerian legal context. The main objectives are as follows:
\begin{itemize}
	\item To investigate the potential of LLMs and Retrieval-Augmented Generation models for processing Arabic legal texts .
	\item To identify and address challenges in document retrieval, particularly the issue of selecting the optimal number of candidate documents (the k-selection problem).
	\item To develop and share a curated dataset of Arabic legal case texts from Algeria, contributing to resources in low-resource legal NLP.
\end{itemize}
\textbf{Contributions :} This thesis makes the following key contributions:
\begin{itemize}
	\item A domain-specific chatbot designed to assist users in navigating Algerian legal case information using natural language queries.
	\item The creation of a new dataset of Arabic-language  legal cases from Algeria, filling a gap in publicly available legal resources for this jurisdiction.
	\item A Dynamic Candidate Selection mechanism, which improves the retrieval phase of RAG by adaptively selecting relevant documents based on contextual thresholds.
\end{itemize}
\textbf{Structure :} 
The thesis is organized into four main chapters: 
\begin{itemize}
	\item Chapter 1: Large Language Models – An overview of LLMs and their capabilities, with a focus on applications in legal NLP.

>>>>

\chapter*{General introduction}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{General introduction}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-MascaraComputer Science: 2025}}}
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
\pagenumbering{arabic}
\addcontentsline{toc}{chapter}{General introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Context :}
In recent years, the legal domain has witnessed a growing interest in the application of advanced natural language processing techniques to improve access to legal information and automate reasoning over juridical data. Legal texts—such as statutes, court decisions, and regulatory documents—are often dense, context-dependent, and linguistically complex, making them challenging to interpret both for non-experts and legal professionals. This complexity is further amplified in multilingual and region-specific legal systems, such as Algeria's\cite{inproceedings}, where access to structured and searchable legal data remains limited.

Large Language Models (LLMs) have emerged as powerful tools for understanding and generating human-like text, showing promise in tasks such as legal question answering, case summarization, and document classification\cite{Naveed2024}. However, their reliance on static training data and the inherent difficulty in capturing fine-grained, up-to-date legal knowledge limits their effectiveness in dynamic and high-stakes domains like law. To bridge this gap, Retrieval-Augmented Generation (RAG) architectures have been proposed, combining the linguistic capabilities of LLMs with external knowledge retrieval systems. This hybrid approach enables models to access and reason over relevant documents at inference time, offering a more grounded and contextually aware output \cite{lewis2020retrieval}.

Although these advances have been made, several challenges persist—particularly in the selection of the most relevant documents from large legal corpora. Poor retrieval quality can significantly degrade the performance and reliability of RAG-based systems. Therefore, refining the retrieval process remains a crucial area of research, especially in low-resource and domain-specific contexts \cite{Information_Retrieval} such as Arabic legal data in Algeria.
\subsubsection*{Problem Statement :}

\begin{itemize}
	\item Large Language Models (LLMs) and Retrieval-Augmented Generation (RAG) architectures offer strong potential for legal AI applications.
	However, their performance is often constrained by weaknesses in the retrieval process.
	\item The selection of irrelevant or loosely related documents can result in hallucinated or misleading responses, which is especially critical in the legal domain.
	\item In the Algerian legal context\cite{inproceedings}, this problem is exacerbated due to:
	\begin{itemize}
		\item The scarcity of structured and standardized legal resources.
		\item The dominance of unstructured legal texts written in Arabic, often lacking annotations or metadata.
	\end{itemize}
\end{itemize}
 These limitations make the retrieval step a key bottleneck in the effectiveness of legal AI systems.
 This thesis aims to address this challenge by developing a more adaptive and context-sensitive retrieval mechanism tailored to the legal domain, specifically focusing on Arabic legal documents.

\subsubsection*{Objectives :} 
This research aims to explore and improve the integration of LLMs and retrieval systems in the legal field, with a focus on the Algerian legal context.   
 The main objectives are as follows:
\begin{itemize}
	\item To investigate the potential of LLMs and Retrieval-Augmented Generation models for processing Arabic legal texts .
	\item To identify and address challenges in document retrieval, particularly the issue of selecting the optimal number of candidate documents (the k-selection problem).
	\item To develop and share a curated dataset of Arabic legal case texts from Algeria, contributing to resources in low-resource legal NLP.
\end{itemize}
\subsubsection*{Contributions :} This thesis makes the following key contributions:
\begin{itemize}
	\item A domain-specific chatbot designed to assist users in navigating Algerian legal case information using natural language queries.
    \item train a complete Retrieval-Augmented Generation (RAG) system fully tailored to Arabic, addressing both the retriever and generator components. This represents one of the first domain-specific implementations of a fully Arabic end-to-end RAG pipeline
	\item The creation of a new dataset of Arabic-language  legal cases from Algeria, filling a gap in publicly available legal resources for this jurisdiction.
	\item A Dynamic Candidate Selection mechanism, which improves the retrieval phase of RAG by adaptively selecting relevant documents based on contextual thresholds.
\end{itemize}
\subsubsection*{Structure :} 
The thesis is organized into four main chapters: 
\begin{itemize}
	\item Chapter 1: Large Language Models – An overview of LLMs and their capabilities, with a focus on applications in legal NLP.
